---
title: "US prison data"
author: "Mitsuo Shiota"
date: "2023-06-07"
format: gfm
execute:
  echo: false
---

Updated: `r Sys.Date()`

## Data source

I have downloaded zip files on annual prison data from [Bureau of Justice Statistics](https://bjs.ojp.gov/library/publications/list?series_filter=Prisoners), extracted, and saved csv files on imprison rates and sentenced prisoners by state in `data` directory.

I have also downloaded some Excel, csv, text files on state resident population from [Census Bureau](https://www.census.gov/data/tables/time-series/demo/popest/2020s-state-total.html), and saved them in `data` directory.

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(readxl)
```

```{r}
#| label: functions
#| include: false

non_state <- c(
      "U", "Federal", "State", "Northeast", "Midwest", "South", "West",
      "District of Columbia", "Dist", "District of Columbia "
    )

na_chars <- c("", "/", "--", ":", "~")

get_year <- function(path) {
  year <- path |> 
    str_remove("data/") |> 
    str_sub(2, 3) |> 
    as.numeric()
  
  if_else(year < 50, year + 2000, year + 1900)
}

read_t07 <- function(path, skip, n_col) {
  year <- get_year(path)
  
  read_csv(path, skip = skip, na = na_chars) |> 
    select(2, 3, all_of(n_col)) |> 
    set_names("state", as.character(year - 1), as.character(year)) |> 
    slice(4:53) |> 
    mutate(state = str_extract(state, "[a-zA-Z ]+"))
}

read_at02 <- function(path, skip, year_col) {
#  year <- get_year(path)
  
  lines <- readLines(path) |>
    str_remove_all("%")
  
read_csv(I(lines), skip = skip, na = na_chars) |> 
    select(2, all_of(year_col)) |> 
    rename(state = ...2) |> 
    filter(!is.na(state)) |> 
    mutate(state = str_extract(state, "[a-zA-Z ]+")) |> 
    filter(!state %in% non_state) |> 
    arrange(state)
}

read_at09 <- function(path, skip, n_col) {
  year <- get_year(path)
  
  read_csv(path, skip = skip, na = na_chars) |> 
    select(2, 3, all_of(n_col)) |> 
    set_names("state", as.character(year - 1), as.character(year)) |> 
    filter(!is.na(state)) |> 
    slice(3:52) |> 
    mutate(state = str_extract(state, "[a-zA-Z ]+")) |> 
    arrange(state)
}

read_at09_1 <- function(path, skip) {
  year <- get_year(path)
  
  read_csv(path, skip = skip) |> 
    select(2, 3) |> 
    set_names("state", as.character(year)) |> 
    filter(!is.na(state)) |> 
    slice(3:52) |> 
    mutate(state = str_extract(state, "[a-zA-Z ]+")) |> 
    arrange(state)
}

read_04 <- function(path, skip) {
  year <- get_year(path)
  
  read_csv(path, skip = skip) |> 
    select(1:4) |> 
    set_names("state", as.character(year), as.character(year - 1), "1995") |> 
    filter(!is.na(state)) |> 
    slice(5:13, 15:26, 28:43, 45:57) |> 
    mutate(state = str_extract(state, "[a-zA-Z ]+")) |> 
    arrange(state)
}

read_04_02 <- function(path, skip, year_col) {
  year <- get_year(path)

  read_csv(path, skip = skip) |> 
    select(1, all_of(year_col)) |> 
    set_names("state", as.character(year), as.character(year - 1)) |> 
    na.omit() |> 
    mutate(state = state |> 
             str_extract("[a-zA-Z ]+") |> 
             str_trim()
           ) |> 
    filter(!state %in% non_state) |> 
    arrange(state)
}
```

## 4 csv files created

```{r}
#| label: imprison_rate
#| message: false
#| warning: false

p21stt07 <- read_t07("data/p21stt07.csv", 12, 7)
p20stt07 <- read_t07("data/p20stt07.csv", 12, 7)
p19t07 <- read_t07("data/p19t07.csv", 12, 9)
p18t07 <- read_t07("data/p18t07.csv", 12, 9)
p17t06 <- read_t07("data/p17t06.csv", 11, 8)
p16t07 <- read_t07("data/p16t07.csv", 12, 8)
p15t06 <- read_t07("data/p15t06.csv", 11, 8)
p14t06 <- read_t07("data/p14t06.csv", 11, 8)
p13t06 <- read_t07("data/p13t06.csv", 12, 8)
p11at03 <- read_t07("data/p11at03.csv", 11, 7)

p10at09 <- read_at09("data/p10at09.csv", 11, 6)
p09at09 <- read_at09("data/p09at09.csv", 11, 7)
p08at10 <- read_at09("data/p08at10.csv", 12, 9)

p07at09 <- read_at09_1("data/p07at09.csv", 11)

imprison_rate <- bind_cols(p08at10[1:2], p09at09[, 2],
                           p10at09[, 2], p11at03[, 2:3],
          p13t06[, 2], p14t06[, 2], p15t06[, 2], p16t07[, 2],
          p17t06[, 2], p18t07[, 2], p19t07[, 2],
          p20stt07[, 2], p21stt07[, 2:3])

write_csv(imprison_rate, "imprison_rate.csv")
```

```{r}
#| label: prisoners
#| message: false
#| warning: false

p21stt02 <- read_t07("data/p21stt02.csv", 11, 6)
p20stt02 <- read_t07("data/p20stt02.csv", 11, 6)
p19t02 <- read_t07("data/p19t02.csv", 11, 7)
p18t02 <- read_t07("data/p18t02.csv", 11, 7)
p17t02 <- read_t07("data/p17t02.csv", 11, 7)
p16t02 <- read_t07("data/p16t02.csv", 12, 7)
p15t02 <- read_t07("data/p15t02.csv", 12, 7)
p14t02 <- read_t07("data/p14t02.csv", 11, 7)
p13t02 <- read_t07("data/p13t02.csv", 12, 7)
p11at01 <- read_at02("data/p11at01.csv", 11, 3:5)
p10at01 <- read_at02("data/p10at01.csv", 11, 3:5)
p09at01 <- read_at02("data/p09at01.csv", 11, 3:5) |> 
  set_names("state", "2000", "2008", "2009")
p08at02 <- read_at02("data/p08at02.csv", 11, c(3, 5, 7)) |> 
  set_names("state", "2000", "2007", "2008")
p07t02 <- read_at02("data/p07t02.csv", 11, 3:5)
p06t01 <- read_at02("data/p06t01.csv", 11, 3:5)|> 
  set_names("state", "2000", "2005", "2006")

p05t03 <- read_04_02("data/p05t03.csv", 12, c(2, 4))
p0403 <- read_04_02("data/p0403.csv", 12, c(2, 4))
p0303 <- read_04_02("data/p0303.csv", 12, c(2, 4))
p0203 <- read_04_02("data/p0203.csv", 12, c(2, 4))
p0103 <- read_04_02("data/p0103.csv", 14, c(2, 4))
p0003p <- read_04_02("data/p0003.csv", 14, c(3, 4))
p9905p <- read_04_02("data/p9905.csv", 16, c(3, 5))
p9803p <- read_04_02("data/P9803.csv", 15, c(2, 3))

prisoners <- bind_cols(p9803p[, c(1, 3)], p9905p[, 3], p0003p[, 3],
          p0103[, 3], p0203[, 3], p0303[, 3],
          p0403[, 3], p05t03[, 3], p06t01[, 3],
          p07t02[, 3], p08at02[, 3:4],
          p11at01[, 2:4], p13t02[, 2:3],
          p15t02[, 2], p16t02[, 2], p17t02[, 2],
          p18t02[, 2], p19t02[, 2], p20stt02[, 2],
          p21stt02[, 2:3])

prisoners <- prisoners |> 
  mutate(state = state.name)

write_csv(prisoners, "prisoners.csv")
```

```{r}
#| label: sentenced_prisoners
#| message: false
#| warning: false

p21stt04 <- read_t07("data/p21stt04.csv", 11, 6)
p20stt04 <- read_t07("data/p20stt04.csv", 11, 6)
p19t04 <- read_t07("data/p19t04.csv", 11, 7)
p18t04 <- read_t07("data/p18t04.csv", 11, 7)
p17t04 <- read_t07("data/p17t04.csv", 11, 7)
p16t04 <- read_t07("data/p16t04.csv", 12, 7)
p15t04 <- read_t07("data/p15t04.csv", 11, 7)
p14t04 <- read_t07("data/p14t04.csv", 11, 7)
p13t04 <- read_t07("data/p13t04.csv", 12, 7)
p11at02 <- read_at02("data/p11at02.csv", 11, 3:6)
p10at04 <- read_at02("data/p10at04.csv", 11, 3:5)
p09at04 <- read_at02("data/p09at04.csv", 11, 3:5) |> 
  set_names("state", "2000", "2008", "2009")
p08at05 <- read_at02("data/p08at05.csv", 11, c(3, 5, 7)) |> 
  set_names("state", "2000", "2007", "2008")
p07at03 <- read_at02("data/p07at03.csv", 11, 3:5)
p06at06 <- read_at02("data/p06at06.csv", 11, 3:5)|> 
  set_names("state", "2000", "2005", "2006")

p05t04 <- read_04("data/p05t04.csv", 11)
p0404 <- read_04("data/p0404.csv", 12)
p0304 <- read_04_02("data/p0304.csv", 12, c(2, 4))
p0204 <- read_04("data/p0204.csv", 12)
p0104 <- read_04_02("data/p0104.csv", 15, c(2, 3))
p0003 <- read_04_02("data/p0003.csv", 14, c(7, 8))
p9905 <- read_04_02("data/p9905.csv", 16, c(9, 11))
p9803 <- read_04_02("data/P9803.csv", 15, c(6, 7))

sentenced_prisoners <- bind_cols(p9803[, c(1, 3)], p9905[, 3], p0003[, 3],
          p0104[, 3], p0204[, 3], p0304[, 3],
          p0404[, 3], p05t04[, 3], p06at06[, 3],
          p07at03[, 3], p08at05[, 3:4],
          p11at02[, 3:5], p13t04[, 2:3],
          p15t04[, 2], p16t04[, 2], p17t04[, 2],
          p18t04[, 2], p19t04[, 2], p20stt04[, 2],
          p21stt04[, 2:3])

sentenced_prisoners <- sentenced_prisoners |> 
  mutate(state = state.name)

write_csv(sentenced_prisoners, "sentenced_prisoners.csv")
```

```{r}
#| label: population
#| message: false
#| warning: false

lines <- readLines("data/st-99-03.txt")

st_94_99 <- read_fwf(I(lines[c(10, 14:78)]), skip = 1) |> 
  select(-1) |> 
  set_names(c("state", "1999", "1998", "1997", "1996", "1995", "1994")) |> 
  slice(15:22, 24:65)

st_90_93 <- read_fwf(I(lines[c(84, 88:152)]), skip = 1) |> 
  select(-1, -7) |> 
  set_names(c("state", "1993", "1992", "1991", "1990")) |> 
  slice(15:22, 24:65)

st_00_09 <- st_est00int_01 <- read_csv("data/st-est00int-01.csv", 
    skip = 3) |> 
  select(-2, -13, -14) |> 
  slice(6:13, 15:56) |> 
  rename(state = ...1) |> 
  mutate(state = str_remove(state, "\\."))

st_10_19 <- read_excel("data/nst-est2020.xlsx",
                       range = "A4:O60") |> 
  select(1, 4:13) |> 
  slice(6:13, 15:56) |> 
  rename(state = ...1) |> 
  mutate(state = str_remove(state, "\\."))

st_20_22 <- read_excel("data/NST-EST2022-POP.xlsx",
                       range = "A4:E60") |> 
  select(-2) |> 
  slice(6:13, 15:56) |> 
  rename(state = ...1) |> 
  mutate(state = str_remove(state, "\\."))

population <- bind_cols(st_90_93, st_94_99[, -1], st_00_09[, -1],
          st_10_19[, -1], st_20_22[, -1])

write_csv(population, "population.csv")
```

```{r}
#| label: year_ranges

imprison_rate_long <- imprison_rate |> 
  pivot_longer(-state, names_to = "year", values_to = "imprison_rate") |> 
  mutate(year = as.numeric(year))

prisoners_long <- prisoners |> 
  pivot_longer(-state, names_to = "year", values_to = "prisoners") |> 
  mutate(year = as.numeric(year)) |> 
  arrange(state, year)

sentenced_prisoners_long <- sentenced_prisoners |> 
  pivot_longer(-state, names_to = "year", values_to = "prisoners") |> 
  mutate(year = as.numeric(year)) |> 
  arrange(state, year)

pop_long <- population |> 
  pivot_longer(-state, names_to = "year", values_to = "pop") |> 
  mutate(year = as.numeric(year)) |> 
  filter(between(year, 1997, 2021)) |> 
  arrange(state, year)

prison_pop <- bind_cols(sentenced_prisoners_long, pop_long[, 3]) |> 
  mutate(imprison_rate = prisoners / pop * 1e5)

yr1 <- range(imprison_rate_long$year)
yr4 <- range(prisoners_long$year)
yr2 <- range(sentenced_prisoners_long$year)
yr3 <- range(pop_long$year)
```

From files in `data` directory, I have made 4 csv files:

- imprison_rate.csv: Number of sentenced prisoners per 100,000 residents by state and year from `r yr1[1]` to `r yr1[2]`,

- prisoners.csv: Number of prisoners by state and year from `r yr4[1]` to `r yr4[2]`,

- sentenced_prisoners.csv: Number of (more than one year) sentenced prisoners by state and year from `r yr2[1]` to `r yr2[2]`, and

- population.csv: Resident population by state and year from `r yr3[1]` to `r yr3[2]`.

## Plot

```{r}
#| label: plot_rates
#| fig-width: 6
#| fig-asp: 0.618
#| out-width: "70%"
#| warning: false

imprison_rate_long |> 
  ggplot(aes(year, imprison_rate)) +
  geom_line(aes(color = state), show.legend = FALSE) +
  labs(x = NULL, y = "Sentenced prisoners per 100K residents",
       title = "Imprison rate",
       subtitle = "from imprison_rate.csv")

prison_pop |> 
  ggplot(aes(year, imprison_rate)) +
  geom_line(aes(color = state), show.legend = FALSE) +
  labs(x = NULL, y = "Sentenced prisoners per 100K residents",
       title = "Imprison rate",
       subtitle = "calculated from sentenced_prisoners.csv and population.csv")
```

```{r}
#| label: plot_numbers
#| fig-width: 6
#| fig-asp: 0.618
#| out-width: "70%"
#| warning: false

prisoners_long |> 
  ggplot(aes(year, prisoners)) +
  geom_line(aes(color = state), show.legend = FALSE) +
  labs(x = NULL, y = "Persons",
       title = "Prisoners",
       subtitle = "from prisoners.csv")

prison_pop |> 
  ggplot(aes(year, prisoners)) +
  geom_line(aes(color = state), show.legend = FALSE) +
  labs(x = NULL, y = "Persons",
       title = "Sentenced prisoners",
       subtitle = "from sentenced_prisoners.csv")

prison_pop |> 
  ggplot(aes(year, pop)) +
  geom_line(aes(color = state), show.legend = FALSE) +
  labs(x = NULL, y = "Persons",
       title = "Resident population",
       subtitle = "from population.csv")
```
